Response to Pachter's Review

### 2020-01-22 Response.Rmd

```{r, warning = FALSE, echo=FALSE, include=FALSE}
linearmodel1 <- lm(df$amscitperyear ~ df$age, data = df)
summary(linearmodel1)
```
```{r, echo = FALSE, include=FALSE}
par(mfrow=c(2,2)) # Change the panel layout to 2 x 2
plot(linearmodel1)
par(mfrow=c(1,1)) # Change back to 1 x 1
```

```{r, echo = FALSE}
plot(df$age,df$amscitperyear, main = "Age vs. MSN Citations per Year Revised Dataset")
abline(linearmodel1, col = "red")
```

```{r, echo=FALSE, include = FALSE}
length(restrictedData$amscitperyear)-sum(is.na(restrictedData$amscitperyear))
```

```{r}
confint(linearmodel1)
```

The slope of the regression line is slightly positive (0.6178, 95% Confidence Interval = (0.115, 1.12)), but the $R^2$ values (Adjusted = 0.01662),  are tragically low. So there is really no correlation. However one could object that we do not have enough data (n = 258), to assess that there is no correlation between citations per year and age. We know this, but thought it would more appropriate to analyze this in a separate paper. However, as noted above, our honor and ability as scientists were attacked so...

### Presenting citations data on every R1 Math Professor with MathSciNet citations
(plus the Institute of Advanced Studies and UC Merced)

```{r, include = FALSE}
allR1 <- read_csv("./FullMathR1Anonymized.csv")
```

```{r, echo = FALSE, include = FALSE}
allR1$Gender <- as.factor(allR1$Gender)
allR1$Role <- as.factor(allR1$Role)
allR1$age <- 2020 - allR1$earliest_pub
allR1<- transform(allR1,  citperyear = citations/age)
summary(allR1)
```

We manually collected the citations and year of first publication of every R1 full math professor by consulting [wikipedia](https://en.wikipedia.org/wiki/List_of_research_universities_in_the_United_States), going to the relevant faculty pages and then collecting MathSciNet citations. We then anonymized it. The 2787 professors we collected data on is in line with [data](http://www.ams.org/profession/data/annual-survey/2016dp-tableDF1.pdf?fbclid=IwAR1mgI0qSEs5nCGquqye741_0lZU-ez7dlcJ3wZYhDtJUswhH1SX7yeiiak) collected by the AMS, after taking into account that about half of universities in the US are classified R2. Great lengths were taken to assess the accuracy of this data, including correlating publications, PhD years, etc. Of course errors in data collection, especially  manual typing errors, happen, but by no means are these errors systematic.

**Exercise 1: Pick your favorite R1 institution, go to MathSciNet, and check how similar our data is to what you determined.**

**Exercise 2: Determine every university without a female professor. We will note that the University of Colorado - Denver has a very strong female professor, but she does not have MathSciNet citations. There should be at least one surprise (and a few non surprises).**

Many aspects of this dataset can, should, and will be analyzed. For now, the following will suffice.

### Citations per Year vs Age
We plot the Age vs Citations per Year for all math R1. We generate a linear regression model and output a 95% confidence interval for the slope. 
```{r, echo=FALSE}
linearmodel2 <- lm(allR1$citperyear ~ allR1$age, data = allR1)
plot(allR1$age, allR1$citperyear, main="Age vs Citations per Year All Math R1")
abline(linearmodel2,col="red") 
```

```{r}
confint(linearmodel2)
```

So while visually it appears that there is no correlation between citations and citations per year, one may object and say, the slope is positive! Which leads to the following question.

**Question: To what power must we raise age to get zero within the confidence interval of slope.**

We object to this question, because the implication of the question is, by how much should we discount the accomplishments of those who are older. Nevertheless, we proceed. 
```{r, echo = FALSE, include=FALSE}
allR1<- transform(allR1,  citperyearPower = citations/age^1.3)
```

```{r, echo = FALSE}
linearmodel3 <- lm(allR1$citperyearPower ~ allR1$age, data = allR1)
plot(allR1$age, allR1$citperyear, main="Age vs Citations per Year^1.3 All Math R1")
abline(linearmodel3,col="red") 
```
```{r}
confint(linearmodel3)
```

It seems raising age to the 1.3 will do the trick.

We will reperform the permutation test comparing citations per year^1.3.

### Citations per Year adjusting for fitted handicap on age

```{r , warning=FALSE, echo = FALSE}
df$amscitperyearpower <- df$amscit/(df$age^1.3)
plot9 <- ggplot(df, aes(x=amscitperyearpower))+
  geom_histogram(data=filter(df, (lettergroup == "A Only"|lettergroup == "A and B")),fill = "red", alpha = 0.2, binwidth = 5) + 
  geom_histogram(data=filter(df, (lettergroup == "B and C"|lettergroup == "B Only"|lettergroup == "A and B")),fill = "blue", alpha = 0.2, binwidth = 5) +
  geom_histogram(data=filter(df, (lettergroup == "C Only"|lettergroup == "B and C")),fill = "green", alpha = 0.2, binwidth = 5) +
  geom_vline(xintercept = mean(filter(df, (lettergroup == "A Only"|lettergroup == "A and B"))$amscitperyearpower, na.rm = TRUE), linetype="dotted", color = "red", size=1.5) +
  geom_vline(xintercept = mean(filter(df, (lettergroup == "B and C"|lettergroup == "B Only"|lettergroup == "A and B"))$amscitperyearpower, na.rm = TRUE), linetype="dotted", color = "blue", size=1.5) +
  geom_vline(xintercept = mean(filter(df, (lettergroup == "C Only"|lettergroup == "B and C"))$amscitperyearpower, na.rm = TRUE), linetype="dotted", color = "green", size=1.5) +
  ggtitle("MathSciNet citperyear^1.3 Comparison (A=red), (B=blue), (C=green)")
plot9
```

```{r, echo = FALSE, include = FALSE}
mean(filter(df, (lettergroup == "A Only"|lettergroup == "A and B"))$amscitperyearpower, na.rm = TRUE)
mean(filter(df, (lettergroup == "B and C"|lettergroup == "B Only"|lettergroup == "A and B"))$amscitperyearpower, na.rm = TRUE)
mean(filter(df, (lettergroup == "C Only"|lettergroup == "B and C"))$amscitperyearpower, na.rm = TRUE)
median(filter(df, (lettergroup == "A Only"|lettergroup == "A and B"))$amscitperyearpower, na.rm = TRUE)
median(filter(df, (lettergroup == "B and C"|lettergroup == "B Only"|lettergroup == "A and B"))$amscitperyearpower, na.rm = TRUE)
median(filter(df, (lettergroup == "C Only"|lettergroup == "B and C"))$amscitperyearpower, na.rm = TRUE)
```

The mean number of citations per year^1.3 for signers of letter A is 6 and the median is 4.
The mean number of citations per year^1.3 for signers of letter B is 14 and the median is 9.
The mean number of citations per year^1.3 for signers of letter C is 19 and the median is 14.

The three hypotheses we would like to assess are:

1. $H_0: \mu(A_{citperyear^{1.3}}) = \mu(B_{citperyear^{1.3}}), H_1: \mu(A_{citperyear^{1.3}}) < \mu(B_{citperyear^{1.3}})$

2. $H_0: \mu(B_{citperyear^{1.3}}) = \mu(C_{citperyear^{1.3}}), H_1: \mu(B_{citperyear^{1.3}}) < \mu(C_{citperyear^{1.3}})$

3. $H_0: \mu(A_{citperyear^{1.3}}) = \mu(C_{citperyear^{1.3}}), H_1: \mu(A_{citperyear^{1.3}}) < \mu(C_{citperyear^{1.3}})$

As above, the vertical red bar is the observed difference for hypothesis 1 in the induced distribution, blue is the observed difference for hypothesis 2, and green is the observed difference for hypothesis 3.

```{r, echo = FALSE}
muA <- mean(filter(df, (lettergroup == "A Only"|lettergroup == "A and B"))$amscitperyearpower, na.rm = TRUE)
muB <- mean(filter(df, (lettergroup == "B and C"|lettergroup == "B Only"|lettergroup == "A and B"))$amscitperyearpower, na.rm = TRUE)
muC <- mean(filter(df, (lettergroup == "C Only"|lettergroup == "B and C"))$amscitperyearpower, na.rm = TRUE)

val1 <- muA - muB
val2 <- muB - muC
val3 <- muA - muC

set.seed(0)
dist <- meanPermutation(na.omit(df$amscitperyearpower),10000)

hist(dist,
     main = "Permutation Test MSN citperyear^1.3 R1 Math Professors",
     xlab = "Differences in Mean")
abline(v=val1, col = "red")
abline(v=val2, col = "blue")
abline(v=val3, col = "green")
```
```{r, echo=FALSE, include = FALSE}
sum(dist < val1)/length(dist)
sum(dist < val2)/length(dist)
sum(dist < val3)/length(dist)
```

The induced p-value for hypothesis 1 is 0. The induced p-value for hypothesis 2 is 0.0207. The induced p-value for hypothesis 3 is 0. Hence we fail to reject hypothesis 2 at a 2% significance level and reject hypotheses 1 and 3 in favor of the alternative. We conclude that after adjusting for age $\mu(A) < \mu(B) \leq \mu(C)$.

## One more check that age is irrelevant when comparing citations

This method was suggested by a friend as a final check to eliminate any question that age was the greatest confounder. We want to show that $\mu(A) < \mu(B \cup C)$. We will randomly sample a population of 20 from A, called $X$. For each member $x \in X$, we will find every person from B and C that is within a two year age interval from $x$. We will randomly sample one, and induce a new population $Y$. Then we will compare the means by storing X-Y. We repeat this 1,000 times and plot a histogram of the induced values. If 0 is within this new distribution, then maybe there is a chance, a totally slim one after above, that in fact age is a confounder. If the distribution is primarily negative, then X < Y. Otherwise X > Y. We perform this analysis with both AMS citations and Google Scholar citations.

```{r, echo=FALSE}
#with amscit
set.seed(0)
df3 <- filter(df, (age>0)&(amscit>0))

ageMatchedRandom <- function(){
  Xp <- filter(df3,lettergroup=="A Only"|lettergroup=="A and B")
  X <- Xp[sample(nrow(Xp),20),]
  Yp <- filter(df3,lettergroup=="B Only"|lettergroup=="A and B"|lettergroup=="B and C"|lettergroup=="C Only")
  #instantiate matrix
  Y <- matrix(0,nrow = 1, ncol = nrow(X))
  #for each person in subsample
  for (i in 1:nrow(X)){
    #get their age
    a <- X[i,]$age
    #create range
    low <- a-2
    high <- a+2
    #find all persons between low and high and then pick one randomly
    agematched <- filter(Yp, (age <= high & age >= low))$amscit
    random <- sample(agematched,1)
    #store
    Y[,i] <- random
  }
  return(mean(X$amscit)-mean(Y))
}

#perform test 1000 times
Dist <- matrix(0,nrow=1,ncol=1000)
for (i in 1:1000){
  Dist[,i]<-ageMatchedRandom()
}
hist(Dist, main="Age Matched Test for MSN Citations")
```

```{r, echo=FALSE, include = FALSE}
sum(Dist>=0)
```

When comparing mathscinet citations with this age matched randomization test, we see that none of the induced distribution is greater than or equal to zero. So when comparing similarly aged apples to apples, $A < B\cup C$

We perform the same analysis with Google Scholar citations.

```{r, echo = FALSE}
#with Google Scholar
set.seed(0)
df3 <- filter(df, (age>0)&(citations>0))

ageMatchedRandom <- function(){
  Xp <- filter(df3,lettergroup=="A Only"|lettergroup=="A and B")
  X <- Xp[sample(nrow(Xp),20),]
  Yp <- filter(df3,lettergroup=="B Only"|lettergroup=="A and B"|lettergroup=="B and C"|lettergroup=="C Only")
  #instantiate matrix
  Y <- matrix(0,nrow = 1, ncol = nrow(X))
  #for each person in subsample
  for (i in 1:nrow(X)){
    #get their age
    a <- X[i,]$age
    #create range
    low <- a-2
    high <- a+2
    #find all persons between low and high and then pick one randomly
    agematched <- filter(Yp, (age <= high & age >= low))$citations
    random <- sample(agematched,1)
    #store
    Y[,i] <- random
  }
  return(mean(X$citations)-mean(Y))
}

#perform test 1000 times
Dist <- matrix(0,nrow=1,ncol=1000)
for (i in 1:1000){
  Dist[,i]<-ageMatchedRandom()
}
hist(Dist, main="Age Matched Test Google Scholar Citations")
```
```{r, echo=FALSE, include = FALSE}
sum(Dist>=0)
```

When comparing Google Scholar citations with this age matched randomization test, we see that 18.1% of the induced distribution is greater than or equal to zero. So when comparing similarly aged apples to apples, it is inconclusive if $A < B\cup C$. Of course, we wonder if this is actually Lior Pachter.

```{r, echo = FALSE}
#with Google Scholar without Pachter
set.seed(0)
df4 <- filter(df, (age>0)&(citations>0)&(name!="lior pachter"))

ageMatchedRandom <- function(){
  Xp <- filter(df4,lettergroup=="A Only"|lettergroup=="A and B")
  X <- Xp[sample(nrow(Xp),20),]
  Yp <- filter(df4,lettergroup=="B Only"|lettergroup=="A and B"|lettergroup=="B and C"|lettergroup=="C Only")
  #instantiate matrix
  Y <- matrix(0,nrow = 1, ncol = nrow(X))
  #for each person in subsample
  for (i in 1:nrow(X)){
    #get their age
    a <- X[i,]$age
    #create range
    low <- a-2
    high <- a+2
    #find all persons between low and high and then pick one randomly
    agematched <- filter(Yp, (age <= high & age >= low))$citations
    random <- sample(agematched,1)
    #store
    Y[,i] <- random
  }
  return(mean(X$citations)-mean(Y))
}

#perform test 1000 times
Dist <- matrix(0,nrow=1,ncol=1000)
for (i in 1:1000){
  Dist[,i]<-ageMatchedRandom()
}
hist(Dist, main="Age Matched Test Google Scholar Citations without Lior Pachter ")
```

```{r, echo=FALSE, include=FALSE}
sum(Dist>=0)
```

When comparing Google Scholar citations, removing Pachter, with this age matched randomization test, we see that 2.7% of the induced distribution is greater than or equal to zero. So when comparing similarly aged apples to apples, it indeed seems that $A < B\cup C$.

## Pachter's Magic Trick: Hypertuning

A note about Pachter's final, "damning," (it is not), figure. He chose a cutoff of age 36, and compared the average Google Scholar citations of letter signers. He finds that if one does this cutoff, the mean citations of A is greater than B. We found this choice of 36 to be curious and somewhat arbitrary. It smelled like parameter tuning, but we wanted to investigate.

We plot the average citations per year and note with a vertical line, the 36 (PhD) age cutoff.

```{r, echo = FALSE}
# Initialize vectors to hold the mean citations of professors younger than a certain age.
x.youngest.age <- 15
x.oldest.age <-65
x.age.range <- x.youngest.age:x.oldest.age
year.span <- x.oldest.age - x.youngest.age
x.a <- vector(mode="numeric", length=year.span)
x.b <- vector(mode="numeric", length=year.span)
x.c <- vector(mode="numeric", length=year.span)

for (x.age in x.age.range)
{
  x.a[1+x.age - x.youngest.age] <- mean(filter(df, ((lettergroup == "A Only"|lettergroup == "A and B")&(role=="professor")&(age <= x.age)))$citations, na.rm = TRUE)
  x.b[1+x.age - x.youngest.age] <- mean(filter(df, ((lettergroup == "B Only"|lettergroup == "A and B"|lettergroup == "B and C")&(role=="professor")&(age <= x.age)))$citations, na.rm = TRUE)
  x.c[1+x.age - x.youngest.age] <- mean(filter(df, ((lettergroup == "C Only"|lettergroup == "B and C")&(role=="professor")&(age <= x.age)))$citations, na.rm = TRUE)
}

plot(x.age.range,x.c,
  type="l",
  col="chartreuse3",
  ylim=c(0,7000),
  main="Average citations of Professors as a function of PhD Age",
  ylab="mean citations",
  xlab="PhD age",
  lwd=2)
#pachter's hyperparameter
abline(v=36)
lines(x.age.range,x.a,
  type="l",
  col="red",
  lwd=2)
legend("topleft",
  c("Letter A","Letter C"),
  fill=c("red","chartreuse3"))
```

```{r, echo=FALSE, include = FALSE}
max(na.omit(filter(df, (citations>0)&(lettergroup=="A Only"|lettergroup=="A and B"))$age))
```
The maximum age since PhD of a letter signer of A is 49. If he were to cutoff his comparison at that point, clearly $C>A$. If he were to cutoff his comparison at 38, $C>A$. Any further left of 36, he would be accused of being biased.

Notice the spike at age 21. This is caused by Lior Pachter. What would happen if we removed Pachter?

```{r, echo = FALSE}
#removing pachter
a.trim <- tail(sort(filter(df, ((lettergroup == "A Only"|lettergroup == "A and B")&(role=="professor")&(age < 36)))$citations),1)
c.trim <- tail(sort(filter(df, ((lettergroup == "C Only"|lettergroup == "B and C")&(role=="professor")&(age < 36)))$citations),1)
df.a.trimmed <- subset(df, ((lettergroup == "A Only"|lettergroup == "A and B")&(role=="professor")&(age < 100)&(!is.na(citations))&!(citations %in% a.trim)))
df.c.trimmed <- subset(df, ((lettergroup == "C Only"|lettergroup == "B and C")&(role=="professor")&(age < 100)&(!is.na(citations))&!(citations %in% c.trim)))

x.a <- 0
x.b <- 0
x.c <- 0

for (x.age in x.age.range)
{
  x.a[1+x.age - x.youngest.age] <- mean(filter(df.a.trimmed, (age <= x.age))$citations, na.rm = TRUE)
  x.c[1+x.age - x.youngest.age] <- mean(filter(df.c.trimmed, (age <= x.age))$citations, na.rm = TRUE)
}

plot(x.age.range,x.c,
  type="l",
  col="chartreuse3",
  ylim=c(0,7000),
  main="Average citations of Professors beneath specified years since PhD \n (Lior Pachter removed)",
  ylab="mean citations",
  xlab="PhD age",
  lwd=2)
lines(x.age.range,x.a,
  type="l",
  col="red",
  lwd=2)
legend("topleft",
  c("Letter A","Letter C"),
  fill=c("red","chartreuse3"))

### 2020-01-22 Response.Rmd

a.trim <- tail(sort(filter(df, ((lettergroup == "A Only"|lettergroup == "A and B")&(role=="professor")&(age < 36)))$citations),2)
c.trim <- tail(sort(filter(df, ((lettergroup == "C Only"|lettergroup == "B and C")&(role=="professor")&(age < 36)))$citations),2)
df.a.trimmed <- subset(df, ((lettergroup == "A Only"|lettergroup == "A and B")&(role=="professor")&(age < 100)&(!is.na(citations))&!(citations %in% a.trim)))
df.c.trimmed <- subset(df, ((lettergroup == "C Only"|lettergroup == "B and C")&(role=="professor")&(age < 100)&(!is.na(citations))&!(citations %in% c.trim)))

```

So it is clear that Pachter's analysis was some sort of magic trick, potentially a thought experiment, and a fraudulent one. It is highly unlikely that a tenured and respected expert in computation and statistics did not know the above result, expecially when a student he suggests take an introductory statistics course immediately spotted it. We claim and assert that he purposefully chose his 36 cutoff to try to undermine our results.

## Tier Rankings

In our excel sheet, (which we understand is the bane of reproducibility), and through the magic of pivot tables, we rank R1 departments by calculating average department citations per year (since first publication).

The top 11 departments using this ranking are:

1. Princeton

2. Institute of Advanced Studies

3. Harvard

4. Stanford

5. University of Chicago

6. University of California - Los Angeles

7. Massachussetts Institute of Technology

8. Columbia University

9. New York University

10. University of Miami

11. University of California - Berkeley

We calculate the average citations/average year since PhD of letters A, B, and C, and compare them to our ranked list.

```{r, echo = FALSE, include = FALSE}
mean(filter(df, ((lettergroup == "A Only"|lettergroup == "A and B")&(role=="professor")&(field=="math")))$amscitperyear, na.rm = TRUE)

mean(filter(df, ((lettergroup == "B and C"|lettergroup == "B Only"|lettergroup == "A and B")&(role=="professor")&(field=="math")))$amscitperyear, na.rm = TRUE)

mean(filter(df, ((lettergroup == "C Only"|lettergroup == "B and C")&(role=="professor")&(field=="math")))$amscitperyear, na.rm = TRUE)
```

The average Math Sci Net Citations per year (PhD Age) is:

1. For letter A - 15.98726

2. For letter B - 41.86467

3. For letter C - 55.3615

Temple has an average citations per year of 12.33, so we retract our claim that letter A is comparable to Temple. It is closer to the University of Massachusetts - Amherst  which has an average citations per year of 16.17. By [US News](https://www.usnews.com/best-graduate-schools/top-science-schools/mathematics-rankings), University of Massachusetts - Amherst's Math Department has a rank of 55. Rutgers has an average citations per year of 35.01, so we retract our claim that letter B is comparable to Rutgers. It is closer to the University of Minesotta which has an average citations per year of 42.07 and a US News Ranking of 19. For Letter C, we claimed that it was another tier higher - indeed it is closer to the University of Chicago, which has an average citations per year of 56.27, ranked 6 by US News.

An astute observer would notice we are not exactly comparing apples to apples. Presumably ones first publication could be before one finishes their PhD. So even with the boost, the order amongst letter signers stands.

## Discussion and Conclusion

We have debunked the claim that age is the greatest contributor for the difference in citations and citations per year between signers of Letter A, B, and C. Indeed, the least meritorious of mathematicians as a whole signed letter A, whereas the more meritorious signed letters B and C, with merit judged by citations. If one was not willing to believe citations impose even a small order on merit, one could replace citations with Fields medals, AMS Fellowships, or many other metrics. One could make extrapolations by coupling this analysis with work by Topaz, but we refrain from doing so.

In this analysis, we have addressed most of the criticisms in Pachter's review, acknowledging our errors when pointed, while rejecting his false claim that age was the greates confounder. The only one we have not addressed is his point that, "several p-values are computed and reported without any multiple testing correction." After consultation with a respected statistician, we do not see what the issue is. We reported every p-value and he is welcome to change the set.seed in our code, which he applauds us as easily reproducible.

Regarding the Russians, one cannot help but make parallels with what is happening now in the USA with what happened in the USSR and China. Mandatory diversity statements are being used as [a proxy for affirmative action](https://whyevolutionistrue.wordpress.com/2019/12/31/life-science-jobs-at-berkeley-with-hiring-giving-precedence-to-diversity-and-inclusion-statements/?fbclid=IwAR1tnpRsEfr43Y4b47EtScpSH_sUdhuqcfjBmi0Ez0uhxrK-aW4lR4C9sfg). Affirmative Action is wrong and is totally pernicious. Advocates of affirmative action seem to believe there is a tradeoff between ability and background, and that someone's race, class, gender, sexual orientation, etc. is an excuse for bad research. This is totally false. There are many strong female mathematicians, there are many strong black mathematicians, there are many strong gay mathematicians. There is nothing about someone's background which can justify bad science.

Pachter ends his review by saying "under-represented minorities and women routinely face discrimination and worse. This is completely unacceptable." We agree. But we do not believe that the solution is to allow for more diverse faculty regardless of their ability as researchers. What we do believe is that increasing opportunities for under represented minorities at an early stage and fairer hiring practices is the solution, not reverse discrimination [advocated by some](https://blogs.ams.org/inclusionexclusion/2017/05/11/get-out-the-way/).

We conclude by reiterating our thanks to Pachter. We truly appreciated your review.

## Data and Code
All code and data used for this report is available at. 
https://github.com/joshp112358/Response-to-Pachter

## References

Lior Pachter's Blog Post - Diversity Matters - January 17, 2020 https://liorpachter.wordpress.com/2020/01/17/diversity-matters/

Chad Topaz's Paper - Version 10 - https://osf.io/preprints/socarxiv/fa4zb/

Our original Paper - Version 1 -  https://arxiv.org/pdf/2001.00670.pdf

In Preparation - A Citation's Analysis of R1 Math Departments by Joshua Paik and Igor Rivin

## Miscellany

There seems to be some squabbles in the comments of Pachter's blog whether the paper is Paik-Rivin or Rivin-Paik. In mathematics, we follow the Hardy-Littlewood rule, namely all authors are first authors and we list authors alphabetically.
